Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2019-05-23 papers/Visual Diagnostics of a Model Explainer/visual-diagnostics-of-a-model-explainer.Rnw
worked on intro to paper

\corres{*Corresponding author. \email{kgoode@iastate.edus}}

@@

Machine learning algorithms have been proven to provide accurate predictions. However, this accuracy comes at a cost. With complicated algorithms, the ability to understand the mechanism in the data and the functional form of the model are lost. This has led for a desire to be able to interpret machine learning algorithms. One approach to remedy this problem would be to develop a model explainer. The objective of such an explainer is to understand and explain model performance.

Locally interpretable model explanations (LIME) is one such model explainer that was developed for this purpose (reference). While some model explainers are focused on understanding a model at the global level, LIME was designed to explain the model for a single prediction of interest. Conceptually, LIME fits a simple interpretable model (the explainer) that is meant to capture the behavior of the complex "black box model" in a local region around the prediction.

@@

\begin{itemize}
\item $x=\left(\begin{array}{cccc}x_1 & x_2 & \hdots & x_p\end{array}\right)\in\mathbb{R}^p$: original representation of an instance being explained (e.g. observations from a set of $p$ continuous features for a specific case in the data) % the LIME paper uses d instead of p for the dimension
\item $x'\in\mathbb{R}^{p'}$: vector for the interpretable representation of the instance being explained (e.g. a vector of indicator variables associated with the features chosen through feature selection indicating whether or not the observed value is in the bin created by LIME for the feature)
\item $G$: class of potentially interpretable models (e.g. linear models, decision trees, rule lists)
\item $g$: explanation model where $g:\{0,1\}^{p'}\rightarrow\mathbb{R}$ and $g\in G$
\item $\Omega(g)$: measure of complexity of $g$ (e.g. depth of a tree, number of non-zero coefficients in a linear model fit using LASSO)
\item $f$: model that is being explained where $f:\mathbb{R}^p\rightarrow\mathbb{R}$ (In classification $f(x)$ is the probability that $x$ belongs to a certain class)
\item $\pi_x(z)$: proximity measure between an instance $z$ to $x$ which defines a locality around $x$
\item $\mathcal{L}(f, g, \Pi_x)$: the fidelity funcions which is a measure of how unfaithful $g$ is in approximating $f$ in the locality defined by $\pi_x$
\item $\xi(x)$: explanation produced by LIME where
  $$\xi(x)=\underset{g\in G}{\arg\min} \ \mathcal{L}(f, g, \Pi_x)+\Omega(g)$$
(i.e. want to minimize $\mathcal{L}(f, g, \Pi_x)$ and keep $\Omega(g)$ low enough to be interpretable by humans)
\end{itemize}

@@

In order to assess the LIME explanations created using different sampling methods, it was of interest to compare the top three features chosen as the important predictors by lime within a case from the test data across the diferent sampling methods. Figure ... is a heatamp showing the top feature chosen by lime for each of the cases in the test data and different bin based sampling methods. The rows represent the cases in the test data, and the columns represent the sampling methods. There are twenty methods included in the plot. These include the equally spaced bins and quantile bins from the lime package and the random forest score tree based bins and the same source tree based bins proposed in this paper. The rows are facetted by the test set and whether or not the observation is a match or not. The columns are faceted by these methods, and the columns within a facet represent the number of bins. Each method has 2 to 6 bins. The colors represent the top feature chosen by lime.

The variables of ccf and cms immediately show up as common variables chosen across all of the sampling methods. However, the patterns across the number of bins withing the sampling methods are different. When equally spaced bins are used, the top feature chosen is consistent across all cases within a number of bins category. For example, ccf is almost always chosen (change to actual number) with 2 equally spaced bins, matches is almost always chosen with 3 equally spaced bins, and non\_cms is always chosen for the nonmatches with 5 and 6 equally spaced bins. This shows that with the bullet matching data, the top feature chosen with equally spaced bins is an artifact of the number of bins used. With equally spaced bins, this figure sugggest that LIME is providing global explanations as opposed to local explanations. It would be preferable that the top feature chosen was more consistent across the number of bins and more variable across the cases. This would suggest that the top feature chosen is dependent on the feature values associated with a particular case and not just on which feature is the best explainer when $b$ number of bins are used.

>>>>

\corres{*Corresponding author. \email{kgoode@iastate.edu}}

@@

Machine learning algorithms are known to provide accurate predictions. (include more about the usefulness of ml algorithms). However, this predictive ability comes at a cost. These algorithms are typically very complex and have no functional form. As a result, the ability to directly interpret the model is lost.

In many areas of application, the need to be able to explain the results from the model is critical. For example, in forensics science a machine learning model could be used to determine the probability that a shoe print found at the crime scene matches the shoe of a suspect. The model may return a high probability of a match, but without the ability to explain which factors in the model led to this high probability, it may be difficult for a jury to trust the results from the model when deciding whether or not to convice the suspect. (add sentence about Europe law) This has led to the desire to be able to interpret machine learning algorithms.

One approach to remedy this problem would be to develop a model explainer that can provide insight into the model performance. A model explainer is a method that is separate from the model but uses the model to provide understanding into mechanism of how the model produces predictions. (try to think of a simple example of a model explainer)

Locally interpretable model explanations (LIME) is one such model explainer (reference). While some model explainers are focused on understanding a model at the global level, LIME was designed to explain the model for a single prediction of interest. Conceptually, LIME fits a simple model that is able to be interpreted, which is referred to as the explainer, that is meant to capture the behavior of the complex "black box model" in a local region around a prediction of interest. (a good place to include an image to help with this explanation)

@@

\begin{itemize}
\item $x=\left(\begin{array}{cccc}x_1 & x_2 & \hdots & x_p\end{array}\right)\in\mathbb{R}^p$: original representation of an instance being explained (e.g. observations from a set of $p$ continuous features for a specific case in the data) % the LIME paper uses d instead of p for the dimension
\item $x'\in\mathbb{R}^{p'}$: vector for the interpretable representation of the instance being explained (e.g. a vector of indicator variables associated with the features chosen through feature selection indicating whether or not the observed value is in the bin created by LIME for the feature)
\item $G$: class of potentially interpretable models (e.g. linear models, decision trees, rule lists)
\item $g$: explanation model where $g:\{0,1\}^{p'}\rightarrow\mathbb{R}$ and $g\in G$
\item $\Omega(g)$: measure of complexity of $g$ (e.g. depth of a tree, number of non-zero coefficients in a linear model fit using LASSO)
\item $f$: model that is being explained where $f:\mathbb{R}^p\rightarrow\mathbb{R}$ (In classification $f(x)$ is the probability that $x$ belongs to a certain class)
\item $\pi_x(z)$: proximity measure between an instance $z$ to $x$ which defines a locality around $x$
\item $\mathcal{L}(f, g, \Pi_x)$: the fidelity functions which is a measure of how unfaithful $g$ is in approximating $f$ in the locality defined by $\pi_x$
\item $\xi(x)$: explanation produced by LIME where
  $$\xi(x)=\underset{g\in G}{\arg\min} \ \mathcal{L}(f, g, \Pi_x)+\Omega(g)$$
(i.e. want to minimize $\mathcal{L}(f, g, \Pi_x)$ and keep $\Omega(g)$ low enough to be interpretable by humans)
\end{itemize}

@@

In order to assess the LIME explanations created using different sampling methods, it was of interest to compare the top three features chosen as the important predictors by lime within a case from the test data across the different sampling methods. Figure ... is a heatamp showing the top feature chosen by lime for each of the cases in the test data and different bin based sampling methods. The rows represent the cases in the test data, and the columns represent the sampling methods. There are twenty methods included in the plot. These include the equally spaced bins and quantile bins from the lime package and the random forest score tree based bins and the same source tree based bins proposed in this paper. The rows are faceted by the test set and whether or not the observation is a match or not. The columns are faceted by these methods, and the columns within a facet represent the number of bins. Each method has 2 to 6 bins. The colors represent the top feature chosen by lime.

The variables of ccf and cms immediately show up as common variables chosen across all of the sampling methods. However, the patterns across the number of bins withing the sampling methods are different. When equally spaced bins are used, the top feature chosen is consistent across all cases within a number of bins category. For example, ccf is almost always chosen (change to actual number) with 2 equally spaced bins, matches is almost always chosen with 3 equally spaced bins, and non\_cms is always chosen for the nonmatches with 5 and 6 equally spaced bins. This shows that with the bullet matching data, the top feature chosen with equally spaced bins is an artifact of the number of bins used. With equally spaced bins, this figure suggest that LIME is providing global explanations as opposed to local explanations. It would be preferable that the top feature chosen was more consistent across the number of bins and more variable across the cases. This would suggest that the top feature chosen is dependent on the feature values associated with a particular case and not just on which feature is the best explainer when $b$ number of bins are used.
